---
title: "Relative Abundance workflow"
author: "JS Pinelo"
date: "2025-04-05"
output: html_document
---
##clean and merge data
We are going to clean and merge the data to make it usable for the graphs and remove any data/rows we dont need We are going to start with the file mothur spits out called "final_opti_mcc.shared" and i loaded it in using whitespace as the delimiter

```{r}
library(tidyverse)
phylatax <- final_opti_mcc %>%
  select(-label, -numOtus)%>%
  rename(sample_id = Group)%>%
  pivot_longer(phylatax,-sample_id, names_to = "otu", values_to = "counts") %>%
  filter(sample_id != "ALGAE") %>% # Taking out algae
  rename_all(tolower)%>%
  mutate(sample_id = recode(sample_id, 
                            "JP_101124_16" = "F24OPR1",
                            "JP_101124_17" = "F24OPR1_2",
                            "JP_101124_18" = "F24OPR1_3",
                            "JP_101124_13"="F24ONR1",
                            "JP_101124_14"="F24ONR1_2",
                            "JP_101124_15"="F24ONR1_3",
                            "JP" = "F24OC1",
                            "JP_101124" = "F24OC1_2",
                            "JP_101124_12" = "F24OC1_3"))#renaming columns with our new naming scheme
```
#filter by site
Sometimes we just want to run certain sites (eg. NR1),we can filter the data for individual sites and rename those chunks for later use

```{r}
nr1phylatax <- filter(phylatax, sample_id %in% c("F24NNR1",
                                                 "F24NNR1_2S",
                                                 "F24NNR1_3S",
                                                 "F24ONR1",
                                                 "F24ONR1_2",
                                                 "F24ONR1_3"))
c1phylatax <- filter(phylatax, sample_id %in% c("F24NC1", "F24NC1_2S", "F24NC1_3S",
                                                "JP" = "F24OC1",
                                                "JP_101124" = "F24OC1_2",
                                                "JP_101124_12" = "F24OC1_3"))
pr1phylatax <- filter(phylatax, sample_id %in% c("F24OPR1", "F24OPR1_2", "F24OPR1_3"))

```
#Load taxonomy
Now that the dataset with OTUs is clean and organized by site we can load in taxonomy, I used "final.opti_mcc.0.03.cons.taxonomy"
```{r}
taxfile <- final_opti_mcc_0_03_cons%>%
  select("OTU","Taxonomy")%>%
  rename_all(tolower)%>%
  mutate(taxfile,taxonomy = str_replace_all(taxonomy,"\\(\\d+\\)",""),
         taxonomy = str_replace(taxonomy, ";$", "")) 
#stringr remove parenthesis (100) with number inside next to taxons 
#"\\(\\d+\\)" is a regular expression argument and basically is (d+) which means a number greater than 1 and 
#replaced it with nothing which is why its just quotes
taxonomy<- separate(taxonomy,
                    col="taxonomy",
                    into = c("kingdom", "phylum", "class", "order", "family", "genus"), 
                    sep = ";")
```
#Join taxonomy and calculate relative abundance
now that we cleaned up the taxonomy file we can join the taxonomy file to whichever site we want and calculate the relative abundances

```{r}
relativeabund<- inner_join(nr1phylatax,taxonomy)%>%
  mutate(rel_abund=counts/sum(counts))%>%
  select(-counts)%>%
  pivot_longer(cols=c("kingdom", "phylum", "class", "order", "family", "genus","otu"), 
               names_to = "level",
               values_to = "taxon")
```
#making a new file just for phyla
```{r}
fallnr1phylaabundance_data <- relativeabund %>%
  filter(level == "phylum")
```
Now we are going to take a look at the data to see what our relative abundances are, in two ways, sum and average.(this doesn't make a figure or any new data, its just for us to see)
#Sum of Relative Abundance (First code) tells you how much of each taxon
is present in each sample, which is useful for understanding the overall 
proportion of a taxon in each sample, but it doesn't account for variation across samples.
```{r}
fallnr1phylaabundance_data %>%
  group_by(sample_id, taxon) %>% 
  summarize(rel_abund=sum(rel_abund),.groups = "drop")%>%
  arrange(desc(rel_abund))
```
#mean relative abundance
Why:gives you the average relative abundance of each taxon, which helps to understand the dominance of taxa in your data, particularly when comparing across samples. Sorting by mean_rel_abund allows you to identify which taxa tend to be more abundant across your dataset.
```{r}
averagefallnr1phyla <-fallnr1phylaabundance_data %>%
  group_by(sample_id, taxon) %>%
  summarize(mean_rel_abund = mean(rel_abund),.groups = "drop")%>%
  arrange(desc(mean_rel_abund))
```
#percentage of each phyla present
Another way we can understand our abundances is to know the percentage of the phyla that makes up the sample

```{r}
percentagenr1 <- fallnr1phylaabundance_data %>%
  filter(level == "phylum") %>%  # Filter for phylum-level taxa
  group_by(sample_id, taxon) %>%  # Group by sample and taxon (phylum)
  summarize(rel_abund = sum(rel_abund), .groups = "drop") %>%  # Sum relative abundance for each taxon in each sample
  group_by(sample_id) %>%  # Group by sample to get the total for each sample
  mutate(perc_abund = rel_abund / sum(rel_abund) * 100) %>%  # Calculate percentage for each phylum in each sample
  arrange(sample_id, desc(perc_abund))  # Optionally arrange by sample_id and percentage

view(Falltop10nr1)
```
#Graphing Average abundance
I am graphing the average abundance cuz its prettier and a little more complicated, this code can be editted down or expanded
```{r}
averagefallnr1phylagraph <- averagefallnr1phyla %>%# Scale the mean relative abundance to ensure it sums to 100% for each sample
  group_by(sample_id) %>%
  mutate(rel_abund_scaled = mean_rel_abund / sum(mean_rel_abund)) %>%
  ggplot(aes(x = sample_id, y = rel_abund_scaled, fill = taxon)) +
  geom_col() + 
  scale_fill_manual(values = phylum_colors)+
  scale_y_continuous(name = "Relative abundance", 
                     labels = scales::percent, 
                     limits = c(0, 1)) + # Set y-axis limits to 0-1 (100%)
  theme(axis.text.x = element_text(angle = 90, hjust = 1))+# Optional: Rotate x-axis labels for readability
  theme_bw() #gets rid of grey background but keeps gridlines

print(averagefallnr1phylagraph)
ggsave("AverageFallnr1PhylaAbundance.png", plot = averagefallnr1phylagraph, 
       width = 12, height = 8, dpi = 300)  
```
#Filtering out phyla
Just for fun, lets filter out bacteria_unclassified and this can be applied to a other phyla. because we edited it within the data file, when we run the graph again it wont have bacteria unclassified
```{r}
fallnr1phylaabundance_data <- fallnr1phylaabundance_data %>%
  group_by(sample_id, taxon) %>% 
  filter(taxon != "Bacteria_unclassified")%>%
  summarize(rel_abund=sum(rel_abund),.groups = "drop")%>%
  arrange(desc(rel_abund))
```








 
